---
title: "BSE3144 - Assessment 2"
author: 
date: "due 2024-05-07 at 9:45 AM"
output:
  pdf_document:
    df_print: tibble
  word_document: default
geometry: margin=1.25in
fontsize: 14
font-family: Garamond
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = 'formatR' )
library(tidyverse)
library(datasets)
```
```{r points, include=FALSE}
# Calculating point totals
assessment <- read_file("Assessment2.Rmd") %>%
  str_extract_all("\\+\\d", simplify = TRUE) %>% 
  str_remove("\\+") %>% as.numeric() 
assessment <- assessment[-1]
```
```{text rubric grading, include=FALSE}
# Rubric/Grading

This project-based assessment contains `r length(assessment)` questions or tasks, with point values varying from `r min(assessment)` to `r max(assessment)`, corresponding to how much effort we expect you to put into each questions or task. Altogether this assessment is worth `r sum(assessment)` points, for reference. So a `r sum(assessment)/10` point question/task should take ~1/10th of the total time you spend on this project. You will have more than 1 whole week in class to finish this project (and the week of spring break to polish it, if you want). We fully expect you to be able to finish this project in the 2+ class hours plus 4-6 hours you should spend outside of class working through materials in a typical week. So that `r sum(assessment)/10` point question/task should take ~30-45 minutes to complete, not including troubleshooting errors. Everyone works at different paces and has different innate abilities, but if you appear to have put much less effort than expected into a question/task you will receive less than the full allotted points. If you have gone above and beyond the requirements of the question/task you may receive more than the allotted points.

Within the document below you will find point totals for each task/question marked with a "+", e.g. +3. 


# Directions

In this assessment we are asking you to choose or identify a data set you are interested in and demonstrate your skills of data wrangling and visualization through asking questions about the data set. 

**This assessment is individual. Everyone must submit a unique document.** However you may work with a partner on the same dataset to allow you to bounce ideas off of each other and discuss your questions and plans to answer them. In general you can feel free to talk to any of your colleagues about what you are working on, as this is known to increase creativity, engagement, and enjoyment, but **everyone must submit an _individual, unique_ document.** 

## Questions/Tasks

In the the "Source" Rmarkdown version of this document, any text that is flanked by two asterisk's `**` (which causes text to appear in bold face in the "Visual" version of the document) is a prompt for you to answer or fill in details below. Outside of code chunks, this text will appear as bold. We have still added the `**` inside of comments in code chunks so you will know where you need to fill in. Basically, in the default Posit theme any blue text is a prompt and black text or code chuncks are for you to replace and fill in. 

## Finding a data set 

You are encouraged to build from your Assessment 1 dataset, but if you would like to explore some different data we have included some recommendations in the Assessment 1 Dataset recommendations section/chunk. 
```
```{text Dataset-recommendations, include = FALSE}
### Dataset recommendations from Assessment 1

Data is everywhere these days. Below we have some recommendations about where to find datasets to use for this assessment, but there are many more options out there, from different governmental agencies, to research papers, to data repositories. Google a topic you are interested in plus "data" and see what you can find, or check out the resources below. 

The critical things a dataset must have for this assessment are:

- ability to be downloaded (as a csv/tsv/excel/sas/stata/spss file ideally)
- at least 1 numerical variable
- at least 1 categorical variable
- something that you care about, so you are willing to put some time into this assessment

#### Health

https://www.who.int/data - The World Health organization has lots of very interesting datasets that are pretty accessible. We would recommend in particular, [Child mortality and causes of death](https://www.who.int/data/gho/data/themes/topics/topic-details/GHO/child-mortality-and-causes-of-death) and [Life expectancy data by country](https://www.who.int/data/gho/data/themes/mortality-and-global-health-estimates/ghe-life-expectancy-and-healthy-life-expectancy).

https://healthdata.gov/ - This site is dedicated to making high value health data more accessible to entrepreneurs, researchers, and policy makers in the hopes of better health outcomes.

https://www.nyam.org/library/collections-and-resources/data-sets/ - The New York Academy of Medicine has links to many publicly available, medical data sets. 

https://www.cdc.gov/datastatistics/index.html - CDC has many datasets, some may be difficult to access or read into R, as they often have their own data visualization tools, but with some digging you can find the raw datasets.

#### Environment

https://waterdata.usgs.gov/nwis - These pages provide access to water-resources data collected at approximately 1.9 million sites across the US and its territories.

https://data.noaa.gov/datasetsearch/ - NOAA has many datasets related to the environment from weather and water, to ecology and environmental health. These are generally pretty accessible too.

#### Agriculture

https://data.nal.usda.gov/ - The USDA has a large collection of agriculturally relevant data sets.

#### General 

https://data.world/datasets - data.world has thousands of data sets on all kinds of different topics that are all open and freely available, but you have to make an account.  

https://datadryad.org/stash - This site is a repository for all kinds of research data, use the search tool to find something you are interested in.

https://udc.vt.edu/ - Virginia Tech's University data commons has all kinds of data about our campus community over time that can be visualized in different ways and downloaded as CSVs. 

https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html - The R datasets package has lots of different data sets on a wide variety of topics

https://www.vdh.virginia.gov/data/ - Virginia Department of Health has lots of datasets, mostly health related but also some environmental datasets. 
```

# Introduction

**Feel free to copy and paste in whatever code and description of your dataset from Assessment 1 here.**

## Questions you could ask and answer using this data

**What are 3 possible questions you could ask about/with this data using the numerical analysis methods we have covered in the last 6 weeks (Regression,Root Finding, Optimization, Systems of Equations, Interpolation/Extrapolation, Integration & Differentiation)?**

1. Question (method)
+2

2. Question (method)
+2

3. Question (method)
+2

## Analysis Plan

**Pick 2 of your above questions that use at least two different numerical methods and write down a step-by-step analysis plan for asking those questions using your chosen dataset. Look back at the many different workflows we have used throughout the semester and replace the example text below with more specifics.**

1. +5 
    a. Download data
    b. Format data
    c. Select relevant data
    d. Plot data
    e. Analyze data
    f. Replot data with analysis 
    

  
2. +5
    a. Download data
    b. Format data
    c. Select relevant data
    d. Plot data
    e. Analyze data
    f. Replot data with analysis
    


# Implementation


**In this section we now want you to demonstrate your ability to wrangle data using functions from the dplyr and tidyr packages, as well as numerical methods functions. Depending on your questions you may not need all of the sections below, if a section is not needed you can just state "this is not applicable" or  Towards answering the questions you pose above complete the following:**

## Question 1 (Method 1)

### Data Wrangling

**Wrangle your data into an appropriate format for asking your question. Describe the steps involved and why they are necessary**
+4

```{r}
# **Replace this with code to wrangle your data**
# +4
```

### Exploratory Plotting
**Make a plot of your data towards answering your questions.**
```{r}
# **Replace this with code to plot your data**
# +4
```

**Describe this plot and how it informs how you will answer your question and the implementation of the numerical method (bounds on optimization, or regression formula for example).**
+4

### Analysis

**Implement your proposed numerical method to answer your question.**
```{r}
# **Replace this with code to analyze your data**
# +4
```

**Are there any other statistical tests that should be done to assess the appropriateness of this numerical method for this dataset?**
+2

**Make a plot of your data with the addition of the results of your numerical analysis.**
```{r}
# **Replace this with code to plot your data**
# +2
```

**Describe the meaning of these results and the certainty of these results.**
+4




## Question 2 (Method 2)

### Data Wrangling

**Wrangle your data into an appropriate format for asking your question. Describe the steps involved and why they are necessary**
+4

```{r}
# **Replace this with code to wrangle your data**
# +4
```

### Exploratory Plotting
**Make a plot of your data towards answering your questions.**
```{r}
# **Replace this with code to plot your data**
# +4
```

**Describe this plot and how it informs how you will answer your question and the implementation of the numerical method (bounds on optimization, or regression formula for example).**
+4

### Analysis

**Implement your proposed numerical method to answer your question.**
```{r}
# **Replace this with code to analyze your data**
# +4
```

**Are there any other statistical tests that should be done to assess the appropriateness of this numerical method for this dataset?**
+2

**Make a plot of your data with the addition of the results of your numerical analysis.**
```{r}
# **Replace this with code to plot your data**
# +2
```

**Describe the meaning of these results and the certainty of these results.**
+4

## Extension

**Find a package, blog, video, or other demo in R that is related in some way to your dataset and questions and introduces functions that we didn't learn about in this class. Provide a link below and briefly summarize what you have found.**
+2

**Write an example, using your dataset, for a function that we didn't learn yet in class. Explain the arguments you use, what the function does, and the output.** 
+6 (3 for example, 3 for explanation)

### Don't forget to submit a reflection on the above to the separate SLO7 assignment.

## Critical Evaluation of AI-use in Data Analysis

**Reflect on the role (real or potential) of AI in your (current or future) data analysis process. Did you use any AI tools (e.g., ChatGPT, Claude, Copilot, automated data cleaning packages, etc.) to assist in brainstorming, code generation, data wrangling, or visualization? If you used these tools,  briefly describe how you used these tools and provide a critical evaluation of the results. What steps did or would you take to verify or validate the output provided by the AI? Discuss the potential benefits, limitations, and ethical considerations of using AI in this context.**
+4

### Don't forget to double check that you are effectively communicating your analysis plan and results, using both writing and plots, to a general audience and submit your whole assessment 2 again to the SLO3 assignment. 